---
title: "Final Project Part 2: COVID-19 Data Analysis"
output:
  pdf_document: default
  html_document: default
date: "2023-06-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

## Covid-19 Historical Data Report

In this report, we will be breaking down COVID data from 2019 until 2023, analyzing COVID-19 case rates and death rates from various regions. Our objective is to answer the following questions:

1. What regions were most or least affected by COVID-19, with respect to cases and deaths?
2. How does the US compare with the rest of the world with respect to death rates as a result of COVID-19?

## Get Data

```{r get_jhu_data}

url_in1 <- 
  "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master"
url_in2 <- 
  "/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_US.csv",
                "time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_US.csv",
                "time_series_covid19_deaths_global.csv")

urls <- str_c(url_in1,url_in2,file_names)

uid_lookup <- 
  "UID_ISO_FIPS_LookUp_Table.csv"
  
uid_lookup_url <- str_c(url_in1,
                        "/csse_covid_19_data/",
                        uid_lookup)
```

## Read Data

```{r import_data, message = FALSE}
global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
us_cases <- read_csv(urls[1])
us_deaths <- read_csv(urls[3])

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
```

\newpage

## Tidy Global Data

We will be tidying the global data. All columns are of interest, we will simply be pivoting the dates to a column in order to get accurate case counts, and appending the case and death documents to each other. We will also be summarizing total cases/deaths per country.

```{r tidy_global_data}
global_cases <- global_cases %>% 
  pivot_longer(cols = -c(`Province/State`,
                         `Country/Region`, Lat, Long),
                          names_to = "date",
                          values_to = "cases")

global_deaths <- global_deaths %>% 
  pivot_longer(cols = -c(`Province/State`,
                         `Country/Region`, Lat, Long),
                          names_to = "date",
                          values_to = "deaths")

global <- global_cases %>%
          full_join(global_deaths) %>%
          rename(Country_Region = `Country/Region`,
                 Province_State = `Province/State`) %>%
          mutate(date = mdy(date))

global <- global %>% filter(cases > 0)

global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID,FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, 
         Population, Combined_Key, Lat, Long) %>%
  rename(region = Country_Region)

global_totals <- global %>% group_by(region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000/Population) %>%
  select(region, date, cases,deaths,deaths_per_mill,Population) %>%
  ungroup()

global_totals <- global_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

global_by_region <- global_totals %>% group_by(region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000/Population) %>%
  select(region, date, cases,deaths,deaths_per_mill,Population) %>%
  ungroup()

global_by_region <- global_by_region %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

global_region_totals <- global_by_region %>%
  group_by(region) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

summary(global_totals)
summary(global_by_region)
```
\newpage 
## Visualizing Regional Data

For a brief visual overview of our data, let's try to reduce the amount of points we look at by looking at a particular region. For example, India.

``` {r plot_global_data}
location <- "India"

global_by_region %>% 
  filter(region == location) %>%
  ggplot(aes(x=date,y=cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = str_c("COVID19 in ", location), y=NULL)


global_by_region %>% 
  filter(region == location) %>%
  ggplot(aes(x=date,y=new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = str_c("COVID19 in ", location), y=NULL)

```
\newpage
## Global Data (Cases)

Let's take a look at the best/worst regions with respect to global COVID-19 cases and deaths. We'll plot the 10 regions that were most/least affected by COVID-19 on a bar chart.

``` {r data_global_cases}
global_region_totals %>% slice_min(cases_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou,everything())

global_region_totals %>% slice_max(cases_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou,everything())

global_region_totals %>%
  top_n(10, cases_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  arrange(cases_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  ggplot(aes(x = region, y = cases_per_thou)) +
  geom_bar(stat = "identity", fill = "red4") +
  coord_flip()

global_region_totals %>%
  top_n(-10, cases_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  arrange(cases_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  ggplot(aes(x = region, y = cases_per_thou)) +
  geom_bar(stat = "identity", fill = "red4") +
  coord_flip()
```
\newpage

## Global Data (Deaths)

``` {r data_global_deaths}
global_region_totals %>% slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou,everything())

global_region_totals %>% slice_max(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou,everything())
  
global_region_totals %>%
  top_n(10, deaths_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  arrange(deaths_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  ggplot(aes(x = region, y = deaths_per_thou)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip()

global_region_totals %>%
  top_n(-10, deaths_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  arrange(deaths_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  ggplot(aes(x = region, y = deaths_per_thou)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip()

```
\newpage

## US Data Tidying

Similar to the Global data, we will tidy the US Data in the same fashion.

``` {r tidy_us_data}
us_cases <- us_cases %>% 
  pivot_longer(cols = -c(UID:Combined_Key),
                          names_to = "date",
                          values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date=mdy(date))


us_deaths <- us_deaths %>% 
  pivot_longer(cols = -c(UID:Population),
                          names_to = "date",
                          values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date=mdy(date))

us <- us_cases %>% full_join(us_deaths)

us <- us %>% filter(cases > 0)

us_by_state <- us %>% group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000/Population) %>%
  select(Province_State, Country_Region, date, cases,deaths,
         deaths_per_mill,Population) %>%
  ungroup()

us_totals <- us_by_state %>% group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000/Population) %>%
  select(Country_Region, date, cases,deaths,deaths_per_mill,Population) %>%
  ungroup()

us_by_state <- us_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
 
us_totals <- us_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

us_state_totals <- us_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)


summary(us_by_state)
summary(us_totals)
```
\newpage

## Visualizing US Data

Let's take a look at total cases across the US and a selected state (New York). We can also look at a heatmap of cases and deaths

``` {r plot_us_data}

us_totals %>% 
  ggplot(aes(x=date,y=cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = "COVID19 in US", y=NULL)

state <- "New York"
us_by_state %>% 
  filter(Province_State == state) %>%
  ggplot(aes(x=date,y=cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = str_c("COVID19 in ", state), y=NULL)

state_map <- map_data("state")

us_state_totals$Province_State <- tolower(us_state_totals$Province_State)
us_state_totals <- rename(us_state_totals, region = Province_State)
state_map <- left_join(state_map, us_state_totals, by = "region")

ggplot(state_map, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = cases_per_thou), color = "grey") + 
  scale_fill_gradient(low = "white", high =  "red")

ggplot(state_map, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = deaths_per_thou), color = "grey") +
  scale_fill_gradient(low = "white", high =  "blue")

```
\newpage

## US New Cases

Let's see how the data looks when we plot the daily new cases.

``` {r us_new_cases}
us_totals %>% 
  ggplot(aes(x=date,y=new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = "COVID19 in US", y=NULL)

state <- "California"
us_by_state %>% 
  filter(Province_State == state) %>%
  ggplot(aes(x=date,y=new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = str_c("COVID19 in ", state), y=NULL)

```
\newpage

## US State Data: Outliers

Like we did for the global data, let's take a look at the states with the best/worst case outcomes with respect to COVID.

``` {r state_outliers}
us_state_totals %>% slice_min(cases_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou,everything())

us_state_totals %>% slice_max(cases_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou,everything())


us_state_totals %>%
  top_n(5, cases_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  arrange(cases_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  ggplot(aes(x = region, y = cases_per_thou)) +
  geom_bar(stat = "identity", fill = "red4") +
  coord_flip()

us_state_totals %>%
  top_n(-5, cases_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  arrange(cases_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  ggplot(aes(x = region, y = cases_per_thou)) +
  geom_bar(stat = "identity", fill = "red4") +
  coord_flip()

us_state_totals %>% slice_min(deaths_per_thou, n=5) %>%
  select(deaths_per_thou, cases_per_thou,everything())

us_state_totals %>% slice_max(deaths_per_thou, n=5) %>%
  select(deaths_per_thou, cases_per_thou,everything())

us_state_totals %>%
  top_n(5, deaths_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  arrange(deaths_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  ggplot(aes(x = region, y = deaths_per_thou)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip()

us_state_totals %>%
  top_n(-5, deaths_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  arrange(deaths_per_thou) %>%
  mutate(region = factor(region, levels=region)) %>%
  ggplot(aes(x = region, y = deaths_per_thou)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip()
```
\newpage

## Cases v. Deaths: United States

One thing we may want to verify is - do cases roughly predict deaths? Our intuition leads us to that conclusion, but lets see how that data checks out for the US data. Note that we've forced a zero intercept - it wouldn't make sense if 0 cases predicted more/less than 0 deaths.

``` {r predictors_state}
mod_state <- lm(deaths_per_thou ~ 0 + cases_per_thou, data = us_state_totals)
summary(mod_state)
x_grid <- seq(1,450)
us_state_totals %>% mutate(pred=predict(mod_state))

us_tot_w_pred <- us_state_totals %>% mutate(pred=predict(mod_state))

us_tot_w_pred %>% ggplot() + 
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```
\newpage

## Cases v. Deaths: Global

We expect a similar result if we are to use the global data as well. We will perform the same linear regression:

``` {r predictors_global}
mod_global <- lm(deaths_per_thou ~ 0 + cases_per_thou, data = global_region_totals)
summary(mod_global)
x_grid <- seq(1,700)
new_df <- tibble(cases_per_thou = x_grid)
global_region_totals %>% mutate(pred=predict(mod_global))

global_tot_w_pred <- global_region_totals %>% mutate(pred=predict(mod_global))

global_tot_w_pred %>% ggplot() + 
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")

```
Notice that the slope for the global data is much higher - we can reasonably conclude that the US was less likely to die from COVID-19. This makes some sense, given that the United States is a first world country and has access to better healthcare than many other countries. However, please take into consideration the bias below.

\newpage

## Bias

For our analysis on the regions most and least affected by COVID-19, we must consider that data could be voluntarily or involuntarily under reported. There may be sociopolitical disadvantages to honestly reporting COVID-19 data which could be Representative in the data. Additionally, regions that are smaller with potentially less capable health infrastructure may be less able to accurately report COVID-19 cases and deaths. One must also be aware of any bias one might have towards particular regions due to culture/race/ethnicity before coming to any conclusions about how a particular region fared with respect to COVID-19.

## Conclusions
Recall the questions we asked at the beginning of the report:

1. What regions were most or least affected by COVID-19, with respect to cases and deaths?
2. How does the US compare with the rest of the world with respect to death rates as a result of COVID-19?

For question 1, we found the following:

### Global Cases/Deaths:
**Most COVID-19 cases per capita:** San Marino, Austria, Slovenia, Brunei, Andorra, Iceland, South Korea, France, Denmark, Liechtenstein

**Least COVID-19 cases per capita:** Sudan, Nigeria, Congo, Burkina Faso, Sierra Leone, Tanzania, Chad, Yemen, Niger, North Korea

**Most COVID-19 deaths per capita:** Peru, Bulgaria, Hungary, Bosnia, North Macedonia, Montenegro, Croatia, Georgia, Czechia, Slovakia.

**Least COVID-19 deaths per capita:** Tanzania, Benin, Tajikistan, Niger, South Sudan, Chad, Burundi, North Korea, Tuvalu, Holy See

### US Cases/Deaths:
**Most COVID-19 cases per capita:** Rhode Island, Alaska, Utah, Kentucky, North Dakota

**Least COVID-19 cases per capita:** Maine, Virgin Islands, Oregon, Maryland, American Samoa

**Most COVID-19 deaths per capita:** Arizona, Oklahoma, Mississippi, West Virginia, New Mexico

**Least COVID-19 deaths per capita:** Vermont, Hawaii, Virgin Islands, Northern Mariana Islands, American Samoa

For question 2, we found that the US fared well compared to the rest of the world in terms of death rates according to our linear model.